---
title: "project"
author: "Franciszek Soldek & Maciej Wrona"
output:
  pdf_document: default
  always_allow_html: true
header-includes:
  - \usepackage{xcolor}
  - \usepackage{float}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{amsfonts}
  - \usepackage{amsthm}
fontsize: 12pt
---


\color{teal}
\centering
# 0. Required packages
\flushleft
\color{black}
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.pos = 'H', cache = TRUE, comment = NA)
knitr::opts_chunk$set(message = FALSE)


if (!requireNamespace("mlr3measures", quietly = TRUE)) install.packages("mlr3measures")
if (!requireNamespace("pdfCluster", quietly = TRUE)) install.packages("pdfCluster")
if (!requireNamespace("clevr", quietly = TRUE)) install.packages("clevr")
if (!requireNamespace("fpc", quietly = TRUE)) install.packages("fpc")
if (!requireNamespace("dendextend", quietly = TRUE)) install.packages("dendextend")
if (!requireNamespace("pdfCluster", quietly = TRUE)) install.packages("pdfCluster")
if (!requireNamespace("poLCA", quietly = TRUE)) install.packages("poLCA")
if (!requireNamespace("clustMD", quietly = TRUE)) install.packages("clustMD")
if (!requireNamespace("ContaminatedMixt", quietly = TRUE)) install.packages("ContaminatedMixt")
if (!requireNamespace("clustvarsel", quietly = TRUE)) install.packages("clustvarsel")
if (!requireNamespace("flexmix", quietly = TRUE)) install.packages("flexmix")
if (!requireNamespace("pgmm", quietly = TRUE)) install.packages("pgmm")
if (!requireNamespace("broom", quietly = TRUE)) install.packages("broom")
if (!requireNamespace("spdep", quietly = TRUE)) install.packages("spdep")
if (!requireNamespace("CARBayes", quietly = TRUE)) install.packages("CARBayes")
if (!requireNamespace("sp", quietly = TRUE)) install.packages("sp")
if (!requireNamespace("tinytex", quietly = TRUE)) install.packages("tinytex")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("cluster", quietly = TRUE)) install.packages("cluster")
```


```{r}
library(mlr3measures)
library(pdfCluster)
library(clevr)
library(fpc)
library(dendextend)
library(poLCA)
library(clustMD)
library(ContaminatedMixt)
library(clustvarsel)
library(flexmix)
library(pgmm)
library(broom)
library(spdep)
library(CARBayes)
library(sp)
library(tinytex)
library(dplyr)
library(cluster)
```

\color{teal}
\centering
# 1. Analysis based on SMR 
\flushleft
\color{black}
```{r cars}

###.csv files

expected_data=read.csv("expected_counts.csv", header=T, sep=",")
observed_data=read.csv("respiratory_admissions.csv", header=T, sep=",")

expected_data$Id=expected_data$code
observed_data$Id=observed_data$IG

expected_data=expected_data[,-1]
observed_data=observed_data[,-1]

```




```{r}
data=merge(expected_data, observed_data, by="Id")
data=as.data.frame(data)
data$SMR2008 <- data$Y2008/data$E2008
data$SMR2009 <- data$Y2009/data$E2009
```

```{r}
###shape files
#install.packages("sf")
library(sf)
shape<-read_sf("SG_IntermediateZoneBdry_2001/")
```

```{r}

sf.data <- merge(shape, data, all.x=FALSE, by.x="IZ_CODE", by.y="Id")
plot(sf.data$geometry)

```

Calculating SMR for each year
```{r}
selected_columns <- c("E2008", "E2009", "Y2008", "Y2009", "SMR2008", "SMR2009")
result_df <- data.frame()

for (col in selected_columns) {
  # Calculate statistics for the current column
  result <- data %>%
    summarise(
      variable = col,
      mean = mean(!!sym(col)),
      median = median(!!sym(col)),
      min = min(!!sym(col)),
      max = max(!!sym(col)),
      sd = sd(!!sym(col)),
      quantile_25 = quantile(!!sym(col), 0.25),
      quantile_50 = quantile(!!sym(col), 0.50),
      quantile_75 = quantile(!!sym(col), 0.75)
    )

  result_df <- rbind(result_df, result)
}

print(result_df)


library(sp)
sp.data<-as_Spatial(sf.data)

```
For the purpose of this analysis, we present two plots, each comprising two maps: one for the year 2008 and another for the year 2009. The second plot is overlaid on OpenStreetMap, which may lead to new insights compared to mapping the risk with ggplots.
```{r, echo=FALSE, results=FALSE}

library(ggplot2)
library(broom)
sp.data@data$id <- rownames(sp.data@data)
temp1<-tidy(sp.data)

#temp1<-st_as_sf(sp.dat)
sp.data2 <- merge(temp1, sp.data@data, by = "id")
sp.data2$SMR2008= sp.data2$Y2008/sp.data2$E2008
sp.data2$SMR2009= sp.data2$Y2009/sp.data2$E2009


library(patchwork)
library(RColorBrewer)
```

Creating the first ggplot for SMR2008
```{r}
# 
plot1 <- ggplot(data = sp.data2, aes(x=long, y=lat, group=group, fill = c(SMR2008))) +
  geom_polygon() +
  coord_equal() +
  xlab("Easting (m)") +
  ylab("Northing (m)") +
  labs(title = "SMR for respiratory hospitalisation, year 2008", fill = "SMR") +
  theme(title = element_text(size=16)) +
  scale_fill_gradientn(colors=brewer.pal(n=9, name="YlOrRd"))
```

Creating the second ggplot for SMR2009
```{r}
# 
plot2 <- ggplot(data = sp.data2, aes(x=long, y=lat, group=group, fill = c(SMR2009))) +
  geom_polygon() +
  coord_equal() +
  xlab("Easting (m)") +
  ylab("Northing (m)") +
  labs(title = "SMR for respiratory hospitalisation, year 2009", fill = "SMR") +
  theme(title = element_text(size=16)) +
  scale_fill_gradientn(colors=brewer.pal(n=9, name="YlOrRd"))

```

Arranging the plots side by side
```{r} 
plots_side_by_side <- plot1 + plot2

plots_side_by_side
```
Creating Leaflet maps
```{r}
library(sp)
sp.dat.ll <- spTransform(sp.data, CRS("+proj=longlat +datum=WGS84 +no_defs"))

library(leaflet)
library(htmltools)
colours_2008 <- colorNumeric(palette = "YlOrRd", domain = sp.dat.ll$SMR2008)
colours_2009 <- colorNumeric(palette = "YlOrRd", domain = sp.dat.ll$SMR2009)

# 
map_2008 <- leaflet(data = sp.dat.ll) %>%
  addTiles() %>%
  addPolygons(fillColor = ~colours_2008(SMR2008),
              color = "", weight = 1,
              fillOpacity = 0.7) %>%
  addLegend(pal = colours_2008, values = ~SMR2008,
            opacity = 1, title = "SMR2008") %>%
  addScaleBar(position = "bottomleft")

map_2009 <- leaflet(data = sp.dat.ll) %>%
  addTiles() %>%
  addPolygons(fillColor = ~colours_2009(SMR2009),
              color = "", weight = 1,
              fillOpacity = 0.7) %>%
  addLegend(pal = colours_2009, values = ~SMR2009,
            opacity = 1, title = "SMR2009") %>%
  addScaleBar(position = "bottomleft")


maps_side_by_side <- tagList(map_2008, map_2009)
maps_side_by_side

```

We also plotted leaflet map (only available in the Markdown file).


The main conclusions are of the visual analisys are as follow:

1. The smaller the civil parish (administrative unit) the higher SMR seems to be. This fact is visible for both years.

2. Civil parishes through which the main roads run seems to have a higher SMR than the other administrative units. However, it would be very difficult to confirm such hypothesis, as we do not have access to such data.

We also present the analysis of spatial autocorrelation in each of the years by using Moran's I statistic.

```{r}
##spatial autocorrelation
library(spdep)
W.nb <- poly2nb(sf.data, row.names = rownames(sf.data))

summary(W.nb)
W <- nb2mat(W.nb, style = "B")

W.list <- nb2listw(W.nb, style = "B")


sf.data$SMR2008= sf.data$Y2008/sf.data$E2008
sf.data$SMR2009= sf.data$Y2009/sf.data$E2009

moran.mc(x = sf.data$SMR2008, listw = W.list, nsim = 10000)
```
We conclude that I statistic equals to 0.404 and is significantly different from independence, thus it provides evidence that there is spatial autocorrelation in the SMR2008 variable
```{r}
moran.mc(x = sf.data$SMR2009, listw = W.list, nsim = 10000)
```
We conclude that I statistic equals to 0.39 and is significantly different from independence, thus it provides evidence that there is spatial autocorrelation in the SMR2009s variable

\color{teal}
\centering
# 2. Leroux model
\flushleft
\color{black}
```{r}
formula2008 <- Y2008 ~ offset(log(E2008)) 
formula2009 <- Y2009 ~ offset(log(E2009))

library(CARBayes)
model2008 <- S.CARleroux(formula=formula2008, family="poisson", data=sf.data, W=W,
burnin=10000, n.sample=100000, thin=10, verbose=FALSE)
print(model2008)

model2009 <- S.CARleroux(formula=formula2009, family="poisson", data=sf.data, W=W,
burnin=10000, n.sample=100000, thin=10, verbose=FALSE)
print(model2009)
```



```{r}
plot_1_2008 <- plot(model2008$samples$rho)
```
The left plot is the traceplot which shows no trend and hence convergence, while the right plot shows a density estimate of the samples. Additionally, these samples show the estimated value of rho ($\rho$) is close to 0.64, suggesting the spatial dependence in these data after adjusting for the covariates is moderate to high.



```{r}
plot_1_2009 <- plot(model2008$fitted.values)
model2008$fitted.values
```
Judging by the plot of SMR2008 mean fitted values, there are no outliers in the data as well as any clear trend. 

```{r}
summary(model2008)
```


These samples show the estimated value of rho ($\rho$) is close to 0.64, suggesting the spatial dependence in these data after adjusting for the covariates is moderate to high.

```{r}
plot(model2009$samples$rho)
```

Judging by the plot of SMR2008 mean fitted values, there are no outliers in the data as well as any clear trend. 
```{r}
plot(model2009$fitted.values)
```

```{r}
summary(model2009)
```

Assessing goodness of fit
```{r}

moran.mc(x = residuals(model2008, type="pearson"), listw = W.list, nsim = 10000)
moran.mc(x = residuals(model2009, type="pearson"), listw = W.list, nsim = 10000)
```

The statistic and accompanying p-value suggest there is no spatial correlation remaining in the residuals from this model, indicating that the spatial CAR model has adequately removed the correlation from the data.


Assessing goodness of fit
```{r}

moran.mc(x = residuals(model2008, type="pearson"), listw = W.list, nsim = 10000)
moran.mc(x = residuals(model2009, type="pearson"), listw = W.list, nsim = 10000)
```

The statistic and accompanying p-value suggest there is no spatial correlation remaining in the residuals from this model, indicating that the spatial CAR model has adequately removed the correlation from the data.

\color{teal}
\centering
# 3. bivariate mixture model
\flushleft
\color{black}
```{r, results=FALSE, echo=FALSE}
########################################
#Clustering simulated data using clustMD
########################################


f.v2008=model2008$fitted.values
f.v2009=model2009$fitted.values
fitted_values=cbind(f.v2008,f.v2009)

#load the clustMD library
#install.packages("clustMD")
library(clustMD)


#Specify variance parameterisations to be used - here all
methods = c("EII", "VII", "EEI", "VEI", "EVI", "VVI")
#Set the range of the number of clusters
clust.range<-c(1:7)


# Initializing storage of the BIC
mds.bic = matrix(NA, nrow = length(clust.range), ncol = length(methods))

# Looping through the 6 variance parameterizations for 1:7 classes (saving BIC) 

for(m in methods){
  for(g in clust.range){
    mds.bic[g, which(methods==m)] = clustMD(fitted_values, G = g, CnsIndx = 2, OrdIndx = 2, Nnorms = 100, MaxIter = 500, model = m, scale = FALSE)$BIChat
  }
}
```


We plot the BIC values for the range of number of clusters with a different line for each variance parameterisation.

```{r}

# Plot the lines
matplot(clust.range, mds.bic, type = "b", pch = 1:6, col = 1:6)
# Add legend
legend("bottomright", legend = c("EII", "VII", "EEI", "VEI", "EVI", "VVI"), col = 1:6, pch = 1:6, ncol = 2)
```
The highest BIC is for "VII" model with 5 clusters. From now on, we will analise mentioned model.    

Now, we fit the model with the highest BIC
```{r}
# 
md.min = clustMD(fitted_values, G = 5, CnsIndx = 2, OrdIndx = 2, Nnorms = 100, MaxIter = 500, model = "VII", scale = FALSE)


```

We plot the data with the clusters labelled by colour.
```{r}
library(MASS)
eqscplot(fitted_values[,1:2], col = md.min$cl, pch = 18)

```

Now, we calculate Average Silhouette Width for the aforementioned model. 
```{r}
d1<-dist(fitted_values)
si1<-silhouette(md.min$cl,d1)
#si1  
ave.silh1<-mean(si1[,3])
ave.silh1

```

\color{teal}
\centering
# 4. k-means algorithm
\flushleft
\color{black}
```{r}


#Check for outliers using single linkage
single.res <- hclust(dist(fitted_values),"single")
plot(single.res)

#Look if cutting tree identifies singletons joining later
temp<-cutree(single.res,k=5)

table(temp)

#Remove the four singleton outliers (only keep observations in clusters 1 or 2 from temp)
new.SMR<-as.matrix(fitted_values[temp==1])


#Run k-means for k from 2 to 10 and record the total within cluster sums of squares for each
res.kmeans.twss<-rep(NA,10)
n<-length(new.SMR)
res.kmeans.twss[1]<-sum((n-1)*apply(new.SMR,2,var))

for(i in 2:10)
{
  res.kmeans.twss[i]<-kmeans(new.SMR,centers=i, nstart=30)$tot.withinss
}

# Plot the elbow graph
plot(c(1:10),res.kmeans.twss,type="b")

# Calculate and plot the gap statistic
gap.kmeans<-clusGap(as.matrix(new.SMR), FUN=kmeans, nstart=30,K.max=10,B=100)
plot(gap.kmeans)
#We see that the bending in the elbow plot is for k=2, so it might suggest that 2 clusters should be considered from now on.


# Calculate the average silhouette width for each k and find the best k
ave.silh<-rep(NA,120)
d<-dist(new.SMR)
length(new.SMR)
for(i in 2:120)
{
  res.kmeans<-kmeans(new.SMR,centers=i, nstart=100)
  si<-silhouette(res.kmeans$cluster,d)
  ave.silh[i]<-mean(si[,3])
}
ave.silh
ave.silh[2]
plot(ave.silh)
#The highest value for ASW is for k=3, suggesting that 3-means is the best fit.

#Compare the 2 cluster k-means solution on the SMR data to the 3 clusters solution

kmeans.2<-kmeans(new.SMR,2,nstart=30)
kmeans.3<-kmeans(new.SMR,3,nstart=30)

#Take a look at the 2-cluster k-medoids clustering and compare it too
library(cluster)
pam.2<-pam(new.SMR,2)
#plot(pam.2)

# Calculate the average silhouette width for each k and find the best k
ave.silh2<-rep(NA,30)
d<-dist(new.SMR)
length(new.SMR)


si2<-silhouette(pam.2$clust,d)
ave.silh2<-mean(si2[,3])

ave.silh2



#cl1<-c(rep(1,109),rep(2,108))
#cl2<-c(rep(1,55),rep(2,54),rep(3,54),rep(4,54))

```


Comparing the results within different algorithms.

```{r}
#ASW for k-means with 3 clusters
ave.silh[2]

#ASW for k-medoids with 3 clusters
ave.silh2

#ASW for Mixture Clustering 
ave.silh1

c(ave.silh[2],ave.silh2,ave.silh1)
```

We see that Average Silhouette Width for 3-means algorithm is the highest, so it is the best algorithm for clustering the mean fitted values SMR data.
